Taking uncertainty in word embedding bias estimation seriously: a bayesian approach

### 2023-05-04 paper/paperDraft.tex
Adds files without models

---
title: 'Taking uncertainty in word embedding bias estimation seriously: a  bayesian
  approach'
author: "Alicja Dobrzeniecka and Rafal Urbaniak"
output:
  pdf_document:
    number_sections: yes
    df_print: kable
    keep_tex: yes
    includes:
      in_header: Rafal_latex7.sty
  html_document:
    df_print: paged
classoption: dvipsnames,enabledeprecatedfontcommands
fontsize: 10pt
documentclass: scrartcl
urlcolor: blue
bibliography: ../references/cosineReferences1.bib
csl: ../references/transactions-on-computational-logic.csl
---

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = '../')

library(ggplot2)
library(ggthemes)
library(gridExtra)
library(dplyr)
library(kableExtra)
library(tidyverse)
library(magrittr)
library(ggpubr)
library(ggExtra)
library(ggthemes)
library(ggforce)
library(latex2exp)
library(Hmisc)
library(grid)

```

# Introduction

Natural language processing (NLP) models are used to perform various language-related tasks such as providing email filters, intelligent assistants, search results,  translations, text analysis and so on. Such tools require input words to be represented by numbers. This can be done with word embeddings, which represent words as vectors of real numbers.  A common way to construct an embedding is to use  a large natural language corpus to train  a  neural network to assign vectors to words that optimize for co-occurrence prediction accuracy. The vectors can then be compared in terms of their similarity -- the usual measure is cosine similarity -- and the results of such comparisons can be used in downstream tasks. Roughly speaking, cosine similarity is an imperfect mathematical proxy for semantic similarity [@Mikolov2013efficient].

It has been suggested [@Bolukbasi2016man; @Caliskan2017semanticsBiases; @Gonen2019lipstick; @Lauscher2019multidimensional; @Garg2018years; @Manzini2019blackToCriminal]  that  such models can learn implicit biases that reflect harmful stereotypical thinking---for example,  the (vector corresponding to the) word \textit{she} might be  much closer in the vector space to  the word \textit{cooking} than  the word \textit{he}. Such phenomena may be undesirable at least  in some  downstream tasks, such as web search, recommendations, and so on. To investigate such issues, several measures of bias in word embeddings have been formulated and applied. Our goal is to use two prominent examples of such measures to argue that this approach oversimplifies the situation and to develop a Bayesian alternative.

One response to the raising of the issue of bias in natural language models might be to say that there is not much point to reflecting on such biases, as they are  unavoidable. This unavoidability might seem in line with the arguments to the effect that learning algorithms are always value-laden [@JohnsonValueFree]: they employ inductive methods that require design-, data-, or risk-related decisions that have to be guided by extra-algorithmic considerations. Such choices necessarily involve value judgments and have to do, for instance, with what simplifications or risks one finds acceptable. 
Admittedly, algorithmic decision making cannot fulfill the value-free ideal, but this only means that even more attention needs to be paid to the values underlying different techniques and decisions, and to the values being pursued in a particular use of an algorithm.

Another response might be to insist that there is no bias introduced by the use of machine learning methods here, since the algorithm is simply learning to correctly predict co-occurrences based on what "reality" looks like. However, this objection overlooks the fact that we, humans, are the ones who construct this linguistic reality, is shaped in part by the natural language processing tools we use on a massive scale. Sure, if there is unfairness and our goal is to diagnose it, we should do complete justice to learning it in the model used in a study thereof. One example of this approach is [@Garg2017hundredYears]. However, if our goal is to develop downstream tools that perform tasks that we care about well without further perpetuating  or exacerbating harmful stereotypes, we still have good reasons to try to minimize the negative impact, at least as long as this does not damage the overall performance of the tool. Moreover, it is often not the case that the corpora mirror reality---to give a trivial example, heads are spoken of more often than kidneys, but this does not mean that kidneys occur much less often in reality than heads. To give a more relevant example, the disproportionate association of female words with female occupations in a corpus actually greatly exaggerates the actual lower disproportion in the real distribution of occupations. [@gordon2012reporting].

In what follows, we focus on  two popular  measures of bias applicable to many existing word embeddings ( \emph{GoogleNews},\footnote{GoogleNews-vectors-negative300, available at  \url{https://github.com/mmihaltz/word2vec-GoogleNews-vectors}.} \emph{Glove}\footnote{Available at \url{https://nlp.stanford.edu/projects/glove/}.} and \emph{Reddit Corpus}\footnote{The Reddit-L2 corpus, available at  \url{http://cl.haifa.ac.il/projects/L2/}.}): \emph{Word Embedding Association Test}  (WEAT) [@Caliskan2017semanticsBiases],  and \emph{Mean Average Cosine Distance} (MAC)  [@Manzini2019blackToCriminal].  We first explain how these measures are supposed to work. Then we argue that they are problematic for various reasons, a key one being that by pre-averaging data they manufacture false confidence, which we illustrate in terms of simulations showing that the measures often suggest the existence of bias even if by design it is non-existent in a simulated dataset.

@@

# Two measures of bias: WEAT and MAC

Conceptually the idea is that if a particular harmful stereotype is learned in a particular embedding, then certain groups of words will be systematically closer to (or further from) each other. This gives rise to the idea of protected groups---for example, in guiding online search completion or recommendation, female words might require protection in that they should not be systematically closer to stereotypically female job names, and male words require protection in that they should not be systematically closer to toxic masculinity stereotypes, such as "tough", "never complaining" or "macho".^[However, for some research-related purposes, such as the study of stereotypes across history [@Garg2017hundredYears], embeddings which do not protect certain classes may also be useful.]

The key role in the measures to be discussed is played by the notion of cosine distance (or, symmetrically, by cosine similarity). These are defined as follows:^[Here, "$-$" stands for point-wise difference, "$\cdot$" stands for the dot product operation, and $\lVert a\rVert  = \sqrt{(a \cdot a)}$.]
\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\lVert  A \rVert \,\lVert B \rVert}
\\
\tag{Distance}
\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B).
\end{align}
Note  that this terminology is slightly misleading, as mathematically cosine distance is not a distance measure, because it does not satisfy the triangle inequality, as generally $\mathsf{cosineDistance}(A,C) \not \leq \mathsf{cosineDistance}(A,B)+ \mathsf{cosineDistance}(B,C)$. We will keep using this mainstream terminology.

One of the first measures of bias has been developed in [@Bolukbasi2016man].  The general idea is that a certain topic is associated with a vector of real numbers (the topic "direction"), and the bias of a word is investigated by considering the projection of its corresponding vector on this direction. For instance, in [@Bolukbasi2016man],   the gender direction $\mathsf{gd}$   is obtained by taking the differences of the vectors corresponding to ten different gendered pairs (such as $\overrightarrow{she} - \overrightarrow{he}$ or $\overrightarrow{girl} - \overrightarrow{boy}$), and then identifying their principal component.^[Roughly, the principal component  is the vector obtained by projecting the data points on their  linear combination in a way that maximizes the variance of the projections.]  The gender bias of a word $w$ is then understood as $w$'s projection on the gender direction: $\vec{w} \cdot gd$ (which, after normalizing by dividing by $\lVert  w \rVert \,\lVert \mathsf{gd} \rVert$, is the same as cosine similarity).  Given a list  $N$ of  supposedly gender neutral words,^[We follow the methodology used in the debate in assuming that there is a class of words  identified as more or less neutral, such as \emph{ballpark, eat, walk, sleep, table}, whose average similarity to the gender direction (or other protected words) is around 0. We will have more to say about this assumption when we describe our dataset construction.]  and the gender direction $\mathsf{gd}$,    the direct gender bias is defined as the average cosine similarity of the words in $N$ from $\mathsf{gd}$ ($c$ is a parameter determining how strict we want to be):
\begin{align*}
\mathsf{directBias_c(N,gd)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},\mathsf{gd})\vert^c}{\vert N \vert }
\end{align*}
\normalsize

The use of projections in bias estimation has been criticized for instance in [@Gonen2019lipstick], where it is pointed out that while a higher average similarity to the  gender direction might be an indicator of bias with respect to a given  class of words, but it is only one possible manifestation of it, and reducing the cosine similarity to such a projection may not be sufficient to eliminate bias. For instance, "math" and "delicate" might be equally similar to a pair of  opposed explicitly gendered words (\emph{she}, \emph{he}), while being closer to quite different stereotypical attribute words (such as \emph{scientific} or \emph{caring}). Further, it is observed in [@Gonen2019lipstick]  that most word pairs retain similarity under debiasing meant to minimize projection-based bias.^[In [@Bolukbasi2016man]  another method which involves analogies and their evaluations by human users on Mechanical Turk is also used. We do not discuss this method in this paper, see its  criticism in [@Nissim2020fair].]

A measure of bias in word embeddings which does not employ similarities to  pre-calculated protected-class-level directions, the Word Embedding Association Test (WEAT), has been proposed in  [@Caliskan2017semanticsBiases]. The idea here is that the   bias between two sets of target words, $X$ and $Y$ (we call them protected words), should be quantified in terms of the cosine similarity  between the protected words and attribute words coming from  two sets of stereotype attribute words, $A$ and $B$ (we will call them attributes). For instance, $X$ might be a set of male names, $Y$ a set of female names, $A$ might contain stereotypically male-related, and $B$ stereotypically female-related career words. The association difference for a particular word $t$ (belonging to either $X$ or $Y$) is:

@@

\begin{align*}
\mathsf{s}(t,A,B) & = \frac{\sum_{a\in A}\mathsf{cos}(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}\mathsf{cos}(t,b)}{\vert B\vert}
\end{align*}
\normalsize
\noindent then, the association difference between $A$ a $B$ is:

@@

\noindent The authors use it as a test statistic in some tests, and the assumption that $X$ and $Y$ are of the same size is necessary for the statistic to make sense. In the final measure (of effect size), WEAT, means are taken, which makes it insensitive to size differences between $X$ and $Y$:
\begin{align}
\mathsf{weat}(A,B) & = \frac{
\mu(\{\mathsf{s}(x,A,B)\}_{x\in X}) -\mu(\{\mathsf{s}(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{\mathsf{s}(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

 WEAT is inspired by  the Implicit Association Test (IAT) [@Nosek2002harvesting] used in psychology, and it uses almost the same word sets, allowing for a \emph{prima facie} sensible comparison with bias in humans. In  [@Caliskan2017semanticsBiases] the authors argue that significant biases---thus measured--- similar to the ones discovered by IAT can be discovered in word embeddings. In [@Lauscher2019multidimensional] the methodology is  extended to a multilingual and cross-lingual setting, arguing that using Euclidean distance instead of cosine similarity does not make much difference, while the bias effects vary greatly across embedding models (interestingly, with social media-text trained embeddings being less biased than those based on Wikipedia).
 A similar methodology is employed in [@Garg2018years]. The authors employ word embeddings trained on corpora from different decades to study the shifts in various biases.^[Strictly speaking, these authors use Euclidean distances and their differences, but the way they take averages and averages thereof is analogous, and so what we will have to say about pre-averaging leading to false confidence applies to this methodology as well.]

<!--  For instance, to compute the occupational embeddings bias for women the authors first compute the average vector of vector embeddings of words that represent women (e.g. \emph{she}, and \emph{female}), then calculate the Euclidean distance between this mean vector and words for occupations. Then they take the mean of these distances and subtract from it the analogously obtained mean for the average vector of vector embeddings of words that represent men. Formally they take the relative norm distance between $X$ and $Y$ to be: -->

<!-- \vspace{-2mm} -->

<!-- \footnotesize -->

<!-- The effect size is computed by normalizing the difference in means as follows: -->

<!-- \begin{align} -->
<!-- \textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2 -->
<!-- \end{align} -->

<!-- \normalsize  -->

<!-- \noindent where the norm used is Euclidean, and $v_X$ and $x_Y$ are average vectors for sets $X$ and $Y$ respectively.  -->
<!-- ] -->

<!-- \vspace{-2mm} -->

WEAT, however, has been developed to investigate biases corresponding to a pair of supposedly opposing stereotypes, and so the question arises as to how generalize the measure to contexts in which biases with respect to more than two stereotypical groups are to be measured. Such a generalization can be found in [@Manzini2019blackToCriminal]. The authors introduce Mean Average Cosine distance (MAC) as a measure of bias. Let $T = \{t_1, \dots, t_k\}$ be a set of protected words, and let each $A_j\in \mathcal{A}$ be a set of attributes stereotypically associated with a protected word where $\mathcal{A}$.
For instance, when biases related to religion are to be investigated, they use a dataset of the  format illustrated in Table \ref{tab:religionOriginal}. The measure is defined as follows:

```{r religionTableHeadEarly,echo=FALSE,eval=FALSE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
religion <- read.csv("datasets/outdated/religionReddit.csv")[-c(1,6,7)]
colnames(religion) <- c("protected words (T)","attributes","atrribute set (A~j)",
                        "cosine distance")

religionOriginal <- religion[!(religion$`atrtibute set (A~j)` %in% c("neutral","human")),]

set.seed(22)
religionSample <- sample_n(religionOriginal, size = 15)
religionSample %>%  kable(format = "latex",booktabs=T,
                          linesep = "",  escape = FALSE, 
                          caption = "Sample 12 rows of the religion dataset.") %>%
  kable_styling(latex_options=c("striped"))
```

@@

\noindent That is, for each  protected word $t\in T$, and each attribute set $A_j$, they first take the mean of distances for this protected word and all attributes in a given attribute class, and then take the mean of thus obtained means for all the protected words and all the protected classes.^[The authors'  code are in their github repository, [https://github.com/TManzini/DebiasMulticlassWordEmbedding](https://github.com/TManzini/DebiasMulticlassWordEmbedding).]

@@

<!-- WEAT will be useful in our criticism of the statistical methods involved, as it is simpler, so the explanation and visualizations will be more transparent (and \emph{mutatis mutandis} this criticism will apply to MAC as well), and MAC will be useful in the development of our alternative method as its range of applicability is the widest. -->

# General methodological problems with cosine-based measures of bias

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  10pt,
  dvipsnames,enabledeprecatedfontcommands]{scrartcl}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
%\documentclass{article}

% %packages
\usepackage{booktabs}
%\usepackage[left]{showlabels}
\usepackage{multirow}
\usepackage{subcaption}

\usepackage{graphicx}
%\usepackage{longtable}
\usepackage{supertabular, booktabs}
\usepackage{ragged2e}
\usepackage{etex}
%\usepackage{yfonts}
\usepackage{marvosym}
\usepackage[notextcomp]{kpfonts}
\usepackage{nicefrac}
\newcommand*{\QED}{\hfill \footnotesize {\sc Q.e.d.}}
\usepackage{floatrow}

\usepackage[textsize=footnotesize]{todonotes}
\newcommand{\ali}[1]{\todo[color=gray!40]{\textbf{Alicja:} #1}}
\newcommand{\mar}[1]{\todo[color=blue!40]{#1}}
\newcommand{\raf}[1]{\todo[color=olive!40]{#1}}

%\linespread{1.5}
\newcommand{\indep}{\!\perp \!\!\! \perp\!}

\setlength{\parindent}{5mm}
\setlength{\parskip}{1pt}

%language
%\usepackage{times}
\usepackage{mathptmx}
\usepackage[scaled=0.86]{helvet}
\usepackage{t1enc}
%\usepackage[utf8x]{inputenc}
%\usepackage[polish]{babel}
%\usepackage{polski}

%AMS
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{mathtools}

\usepackage{geometry}
 \geometry{a4paper,left=35mm,top=20mm,}

%environments
\newtheorem{fact}{Fact}

%abbreviations
\newcommand{\ra}{\rangle}
\newcommand{\la}{\langle}
\newcommand{\n}{\neg}
\newcommand{\et}{\wedge}
\newcommand{\jt}{\rightarrow}
\newcommand{\ko}[1]{\forall  #1\,}
\newcommand{\ro}{\leftrightarrow}
\newcommand{\exi}[1]{\exists\, {_{#1}}}
\newcommand{\pr}[1]{\ensuremath{\mathsf{P}(#1)}}
\newcommand{\cost}{\mathsf{cost}}
\newcommand{\benefit}{\mathsf{benefit}}
\newcommand{\ut}{\mathsf{ut}}

\newcommand{\odds}{\mathsf{Odds}}
\newcommand{\ind}{\mathsf{Ind}}
\newcommand{\nf}[2]{\nicefrac{#1\,}{#2}}
\newcommand{\R}[1]{\texttt{#1}}
\newcommand{\prr}[1]{\mbox{$\mathtt{P}_{prior}(#1)$}}
\newcommand{\prp}[1]{\mbox{$\mathtt{P}_{posterior}(#1)$}}

\newtheorem{q}{\color{blue}Question}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}[fact]

%technical intermezzo
%---------------------

\newcommand{\intermezzoa}{
	\begin{minipage}[c]{13cm}
	\begin{center}\rule{10cm}{0.4pt}

	\tiny{\sc Optional Content Starts}

	\vspace{-1mm}

	\rule{10cm}{0.4pt}\end{center}
	\end{minipage}\nopagebreak 
	}

\newcommand{\intermezzob}{\nopagebreak 
	\begin{minipage}[c]{13cm}
	\begin{center}\rule{10cm}{0.4pt}

	\tiny{\sc Optional Content Ends}

	\vspace{-1mm}

	\rule{10cm}{0.4pt}\end{center}
	\end{minipage}
	}

%--------------------

\newtheorem*{reply*}{Reply}
\usepackage{enumitem}
\newcommand{\question}[1]{\begin{enumerate}[resume,leftmargin=0cm,labelsep=0cm,align=left]
\item #1
\end{enumerate}}

\usepackage{float}

\usepackage{sectsty}
\sectionfont{\fontsize{11}{11}\selectfont}

% \setbeamertemplate{blocks}[rounded][shadow=true]
% \setbeamertemplate{itemize items}[ball]
% \AtBeginPart{}
% \AtBeginSection{}
% \AtBeginSubsection{}
% \AtBeginSubsubsection{}
% \setlength{\emergencystretch}{0em}
% \setlength{\parskip}{0pt}

\usepackage[authoryear]{natbib}

%\bibliographystyle{apalike}

\usepackage{tikz}
\usetikzlibrary{positioning,shapes,arrows}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Taking uncertainty in word embedding bias estimation seriously: a bayesian approach},
  pdfauthor={Alicja Dobrzeniecka and Rafal Urbaniak},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

\title{Taking uncertainty in word embedding bias estimation seriously: a
bayesian approach}
\author{Alicja Dobrzeniecka and Rafal Urbaniak}
\date{}

\begin{document}
\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Natural language processing (NLP) models are used to perform various
language-related tasks such as providing email filters, intelligent
assistants, search results, translations, text analysis and so on. Such
tools require input words to be represented by numbers. This can be done
with word embeddings, which represent words as vectors of real numbers.
A common way to construct an embedding is to use a large natural
language corpus to train a neural network to assign vectors to words
that optimize for co-occurrence prediction accuracy. The vectors can
then be compared in terms of their similarity -- the usual measure is
cosine similarity -- and the results of such comparisons can be used in
downstream tasks. Roughly speaking, cosine similarity is an imperfect
mathematical proxy for semantic similarity {[}14{]}.

It has been suggested {[}1,4,5,8,11,12{]} that such models can learn
implicit biases that reflect harmful stereotypical thinking---for
example, the (vector corresponding to the) word \textit{she} might be
much closer in the vector space to the word \textit{cooking} than the
word \textit{he}. Such phenomena may be undesirable at least in some
downstream tasks, such as web search, recommendations, and so on. To
investigate such issues, several measures of bias in word embeddings
have been formulated and applied. Our goal is to use two prominent
examples of such measures to argue that this approach oversimplifies the
situation and to develop a Bayesian alternative.

One response to the raising of the issue of bias in natural language
models might be to say that there is not much point to reflecting on
such biases, as they are unavoidable. This unavoidability might seem in
line with the arguments to the effect that learning algorithms are
always value-laden {[}9{]}: they employ inductive methods that require
design-, data-, or risk-related decisions that have to be guided by
extra-algorithmic considerations. Such choices necessarily involve value
judgments and have to do, for instance, with what simplifications or
risks one finds acceptable. Admittedly, algorithmic decision making
cannot fulfill the value-free ideal, but this only means that even more
attention needs to be paid to the values underlying different techniques
and decisions, and to the values being pursued in a particular use of an
algorithm.

Another response might be to insist that there is no bias introduced by
the use of machine learning methods here, since the algorithm is simply
learning to correctly predict co-occurrences based on what ``reality''
looks like. However, this objection overlooks the fact that we, humans,
are the ones who construct this linguistic reality, is shaped in part by
the natural language processing tools we use on a massive scale. Sure,
if there is unfairness and our goal is to diagnose it, we should do
complete justice to learning it in the model used in a study thereof.
One example of this approach is {[}3{]}. However, if our goal is to
develop downstream tools that perform tasks that we care about well
without further perpetuating or exacerbating harmful stereotypes, we
still have good reasons to try to minimize the negative impact, at least
as long as this does not damage the overall performance of the tool.
Moreover, it is often not the case that the corpora mirror reality---to
give a trivial example, heads are spoken of more often than kidneys, but
this does not mean that kidneys occur much less often in reality than
heads. To give a more relevant example, the disproportionate association
of female words with female occupations in a corpus actually greatly
exaggerates the actual lower disproportion in the real distribution of
occupations. {[}6{]}.

In what follows, we focus on two popular measures of bias applicable to
many existing word embeddings (
\emph{GoogleNews},\footnote{GoogleNews-vectors-negative300, available at  \url{https://github.com/mmihaltz/word2vec-GoogleNews-vectors}.}
\emph{Glove}\footnote{Available at \url{https://nlp.stanford.edu/projects/glove/}.}
and
\emph{Reddit Corpus}\footnote{The Reddit-L2 corpus, available at  \url{http://cl.haifa.ac.il/projects/L2/}.}):
\emph{Word Embedding Association Test} (WEAT) {[}8{]}, and
\emph{Mean Average Cosine Distance} (MAC) {[}12{]}. We first explain how
these measures are supposed to work. Then we argue that they are
problematic for various reasons, a key one being that by pre-averaging
data they manufacture false confidence, which we illustrate in terms of
simulations showing that the measures often suggest the existence of
bias even if by design it is non-existent in a simulated dataset.

@@

\hypertarget{two-measures-of-bias-weat-and-mac}{%
\section{Two measures of bias: WEAT and
MAC}\label{two-measures-of-bias-weat-and-mac}}

Conceptually the idea is that if a particular harmful stereotype is
learned in a particular embedding, then certain groups of words will be
systematically closer to (or further from) each other. This gives rise
to the idea of protected groups---for example, in guiding online search
completion or recommendation, female words might require protection in
that they should not be systematically closer to stereotypically female
job names, and male words require protection in that they should not be
systematically closer to toxic masculinity stereotypes, such as
``tough'', ``never complaining'' or ``macho''.\footnote{However, for
  some research-related purposes, such as the study of stereotypes
  across history {[}3{]}, embeddings which do not protect certain
  classes may also be useful.}

The key role in the measures to be discussed is played by the notion of
cosine distance (or, symmetrically, by cosine similarity). These are
defined as follows:\footnote{Here, ``\(-\)'' stands for point-wise
  difference, ``\(\cdot\)'' stands for the dot product operation, and
  \(\lVert a\rVert = \sqrt{(a \cdot a)}\).} \begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\lVert  A \rVert \,\lVert B \rVert}
\\
\tag{Distance}
\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B).
\end{align} Note that this terminology is slightly misleading, as
mathematically cosine distance is not a distance measure, because it
does not satisfy the triangle inequality, as generally
\(\mathsf{cosineDistance}(A,C) \not \leq \mathsf{cosineDistance}(A,B)+ \mathsf{cosineDistance}(B,C)\).
We will keep using this mainstream terminology.

One of the first measures of bias has been developed in {[}1{]}. The
general idea is that a certain topic is associated with a vector of real
numbers (the topic ``direction''), and the bias of a word is
investigated by considering the projection of its corresponding vector
on this direction. For instance, in {[}1{]}, the gender direction
\(\mathsf{gd}\) is obtained by taking the differences of the vectors
corresponding to ten different gendered pairs (such as
\(\overrightarrow{she} - \overrightarrow{he}\) or
\(\overrightarrow{girl} - \overrightarrow{boy}\)), and then identifying
their principal component.\footnote{Roughly, the principal component is
  the vector obtained by projecting the data points on their linear
  combination in a way that maximizes the variance of the projections.}
The gender bias of a word \(w\) is then understood as \(w\)'s projection
on the gender direction: \(\vec{w} \cdot gd\) (which, after normalizing
by dividing by \(\lVert w \rVert \,\lVert \mathsf{gd} \rVert\), is the
same as cosine similarity). Given a list \(N\) of supposedly gender
neutral words,\footnote{We follow the methodology used in the debate in
  assuming that there is a class of words identified as more or less
  neutral, such as \emph{ballpark, eat, walk, sleep, table}, whose
  average similarity to the gender direction (or other protected words)
  is around 0. We will have more to say about this assumption when we
  describe our dataset construction.} and the gender direction
\(\mathsf{gd}\), the direct gender bias is defined as the average cosine
similarity of the words in \(N\) from \(\mathsf{gd}\) (\(c\) is a
parameter determining how strict we want to be): \begin{align*}
\mathsf{directBias_c(N,gd)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},\mathsf{gd})\vert^c}{\vert N \vert }
\end{align*} \normalsize

The use of projections in bias estimation has been criticized for
instance in {[}5{]}, where it is pointed out that while a higher average
similarity to the gender direction might be an indicator of bias with
respect to a given class of words, but it is only one possible
manifestation of it, and reducing the cosine similarity to such a
projection may not be sufficient to eliminate bias. For instance,
``math'' and ``delicate'' might be equally similar to a pair of opposed
explicitly gendered words (\emph{she}, \emph{he}), while being closer to
quite different stereotypical attribute words (such as \emph{scientific}
or \emph{caring}). Further, it is observed in {[}5{]} that most word
pairs retain similarity under debiasing meant to minimize
projection-based bias.\footnote{In {[}1{]} another method which involves
  analogies and their evaluations by human users on Mechanical Turk is
  also used. We do not discuss this method in this paper, see its
  criticism in {[}16{]}.}

A measure of bias in word embeddings which does not employ similarities
to pre-calculated protected-class-level directions, the Word Embedding
Association Test (WEAT), has been proposed in {[}8{]}. The idea here is
that the bias between two sets of target words, \(X\) and \(Y\) (we call
them protected words), should be quantified in terms of the cosine
similarity between the protected words and attribute words coming from
two sets of stereotype attribute words, \(A\) and \(B\) (we will call
them attributes). For instance, \(X\) might be a set of male names,
\(Y\) a set of female names, \(A\) might contain stereotypically
male-related, and \(B\) stereotypically female-related career words. The
association difference for a particular word \(t\) (belonging to either
\(X\) or \(Y\)) is:

@@

\begin{align*}
\mathsf{s}(t,A,B) & = \frac{\sum_{a\in A}\mathsf{cos}(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}\mathsf{cos}(t,b)}{\vert B\vert}
\end{align*} \normalsize \noindent then, the association difference
between \(A\) a \(B\) is:

@@

\noindent The authors use it as a test statistic in some tests, and the
assumption that \(X\) and \(Y\) are of the same size is necessary for
the statistic to make sense. In the final measure (of effect size),
WEAT, means are taken, which makes it insensitive to size differences
between \(X\) and \(Y\): \begin{align}
\mathsf{weat}(A,B) & = \frac{
\mu(\{\mathsf{s}(x,A,B)\}_{x\in X}) -\mu(\{\mathsf{s}(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{\mathsf{s}(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

WEAT is inspired by the Implicit Association Test (IAT) {[}17{]} used in
psychology, and it uses almost the same word sets, allowing for a
\emph{prima facie} sensible comparison with bias in humans. In {[}8{]}
the authors argue that significant biases---thus measured--- similar to
the ones discovered by IAT can be discovered in word embeddings. In
{[}11{]} the methodology is extended to a multilingual and cross-lingual
setting, arguing that using Euclidean distance instead of cosine
similarity does not make much difference, while the bias effects vary
greatly across embedding models (interestingly, with social media-text
trained embeddings being less biased than those based on Wikipedia). A
similar methodology is employed in {[}4{]}. The authors employ word
embeddings trained on corpora from different decades to study the shifts
in various biases.\footnote{Strictly speaking, these authors use
  Euclidean distances and their differences, but the way they take
  averages and averages thereof is analogous, and so what we will have
  to say about pre-averaging leading to false confidence applies to this
  methodology as well.}

WEAT, however, has been developed to investigate biases corresponding to
a pair of supposedly opposing stereotypes, and so the question arises as
to how generalize the measure to contexts in which biases with respect
to more than two stereotypical groups are to be measured. Such a
generalization can be found in {[}12{]}. The authors introduce Mean
Average Cosine distance (MAC) as a measure of bias. Let
\(T = \{t_1, \dots, t_k\}\) be a set of protected words, and let each
\(A_j\in \mathcal{A}\) be a set of attributes stereotypically associated
with a protected word where \(\mathcal{A}\). For instance, when biases
related to religion are to be investigated, they use a dataset of the
format illustrated in Table \ref{tab:religionOriginal}. The measure is
defined as follows:

@@

\noindent That is, for each protected word \(t\in T\), and each
attribute set \(A_j\), they first take the mean of distances for this
protected word and all attributes in a given attribute class, and then
take the mean of thus obtained means for all the protected words and all
the protected classes.\footnote{The authors' code are in their github
  repository,
  \url{https://github.com/TManzini/DebiasMulticlassWordEmbedding}.}

@@

\hypertarget{general-methodological-problems-with-cosine-based-measures-of-bias}{%
\section{General methodological problems with cosine-based measures of
bias}\label{general-methodological-problems-with-cosine-based-measures-of-bias}}
